On the formation of H$\alpha$ line emission around classical T Tauri stars

### 2006-01-18 TTau_Halpha/kurosawa.tex
end of wednesday  -th
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@620 25464f21-6099-402c-b30a-640f1d40d38b

We present  radiative transfer models of classical T~Tauri stars, and
investigate the formation of H$\alpha$ line emission 
in order to understand their complex circumstellar environments, and to explain
the wide variety of line profiles seen in observations. To
overcome the limited applicability of the radiative transfer models
with magnetospherical accretion alone to fitting observed H$\alpha$
profiles, and to any other lines affected by the stellar wind,
two types of kinematic wind models are introduced: (1)~the bipolar
wind model in which the wind is originated from the star itself, and
(2)~the disc-wind model in which the wind originates from the inner
part of the accretion disc. We perform systematic investigations of
the model parameters for the wind and the magnetosphere to search
for possible geometrical and physical conditions which lead to the
types of profiles seen in observations. We find both wind models
can reproduce the wide range profile types seen is observations, 
although the inclination dependence of the line equivalent width
predicted by the bipolar wind model agree with trends seen in the
observation, but the disc-wind model does not. Using the model results,
we examine the H$\alpha$ spectroscopic classification used by Reipurth
et.~al, and discuss the basic physical conditions required to reproduce
the profiles in each classified type.

@@

Many observational studies (e.g., \citealt{herbig:1962}; \citealt{edwards:1994};
\citealt{kenyon:1994}; \citealt*{reipurth:1996}; \citealt{alencar:2000})
of CTTS line profiles show evidence for both outward wind flows and
inward accretion flows, as seen in the blue-shifted absorption features
in H$\alpha$ profiles and the redshifted inverse P~Cygni (IPC) profiles
respectively. Typical mass-loss rates of CTTS are about $10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$
to $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$ (e.g., \citealt{kuhi:1964};
\citealt{edwards:1987}; \citealt*{hartigan:1995}), and the mass-accretion
rates are also about $10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$ to $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$
(e.g., \citealt{kenyon:1987}; \citealt*{bertout:1988}; \citealt{gullbring:1998}).
Recent H$\alpha$ spectro-astrometric observations by \citet*{takami:2003}
show direct evidence for the presence of bipolar and monopolar
outflows down to $\sim 1\,\mathrm{au}$ scale (e.g.\,CS~Cha and RU~Lup).
Similarly, ESO VLT observations using high-resolution ($R=50\,000$)
two-dimensional spectra of edge-on CTTS (HH30$^{*}$, HK~Tau~B,
and HV~Tau~C) by \citet{appenzeller:2005} show the extended H$\alpha$
emission in the direction perpendicular to the obscuring circumstellar
disc both above and below the disc, suggesting the presence
of the bipolar outflows. On a larger scale, \emph{HST} observations
of HH30 \citep{burrows:1996} show the jet traced to within $\lesssim30$~au
from the star. The jet has a cone shape with an opening angle of $3^{\circ}$
between 70 and 700 au \citep{koenigl:2000}. \citet{alencar:2000}
found about 80 per cent of their sample (30 CTTS) show blue-shifted
absorption components in at least one of the Balmer lines and Ca~K (most
commonly in H$\alpha$).

In the currently favoured model of accretion in CTTS, the
accretion disc is disrupted by the magnetosphere of stars which
channels the gas from the disc onto the stellar surface (e.g.,
\citealt{uchida:1985}; \citealt{koenigl:1991}; \citealt{cameron:1993};
\citealt{shu:1994}). This picture is supported
by the evidence that CTTS have relatively strong ($\sim 10^{3}\mathrm{G}$)
magnetic field (e.g., \citealt{johns-krull:1999};
\citealt{guenther:1996};  \citealt{symington:2005b})
and by radiative transfer models which reproduce the observed
profiles for some TTS (\citealt*{muzerolle:2001}). The magnetospherical
accretion model naturally explains the blue-ward asymmetric emission
lines (seen in some of CTTS) caused by the partial occultation of
the flow by the stellar disc, and the redshifted absorption component at the
typical free-fall velocities (a few hundred $\mathrm{km\, s^{-1}}$)
seen in some of CTTS. Despite the success of the magnetospherical
accretion model in explaining the line profiles in some CTTS, the
overwhelming observational evidence for outflow (mentioned above)
in the CTTS profiles suggests that this model is only a part of the
complete picture. Clearly, a modification to include an out-flowing
wind/jet flow is necessary if one wishes to predict the mass-accretion
rate and the mass-loss rate of CTTS by modelling their emission profiles
(e.g. H$\alpha$).

Prior to the magnetospherical paradigm, many alternative models had
been considered to explain the observed spectroscopic features mentioned
earlier. For example, (1)~the Alfv\'{e}n wave-driven wind model
(e.g.~\citealt{decampli:1981}; \citealt*{hartmann:1982}), (2)~turbulent
boundary layer (between the accretion disc and stellar surface) model
(e.g.~\citealt{bertout:1988}; \citealt{basri:1989}), (3)~chromospheric
model (e.g.~\citealt*{calvet:1984}), and (4)~disc wind model (e.g.
\citealt*{calvet:1992b}; \citealt{kwan:1995}).  Recent work by
\citet{alencar:2005} showed the observed 
H$\alpha$, H$\beta$ and Na~D lines of RW Aur are better be produced
by the radiative transfer model which includes a collimated disc-wind
arising from near the inner edge of the accretion disc. \citet{malbet:2005}
considered a combination of a spherical wind and an accretion disc
in their radiative transfer model to reproduce Br$\gamma$ from the
early-type Herbig Be star MWC~297.

The magneto-centrifugal wind model, first proposed by \citet{blandford:1982},
has been often used to reproduce the large-scale wind structure of
T Tauri stars, or to model observed optical jets (e.g.~HH~30 jet
by \citealt{burrows:1996}; \citealt{ray:1996}). The launching of
the wind from a Keplerian disc is typically done by treating the equatorial
plane of the disc as a mass-injecting boundary condition (e.g., \citealt{shu:1994};
\citealt{ustyugova:1995}; \citealt{ouyed:1997}; \citealt{krasnopolsky:2003}).
Depending on the location of the open magnetic fields anchored to
the disc, two different types of winds are produced. If the field
is constrained to be near the co-rotation radius of stellar magnetosphere,
an {}``X-wind'' \citep{shu:1994} is produces. If the open field
lines are located in a wider area of the disc, a {}``disc-wind''
similar to \citet{koenigl:2000} is produced \citet{krasnopolsky:2003}.
Recent reviews on the jet/wind-disc connection can be found in \citet{koenigl:2000}
and \citet{pudritz:2005}. Interestingly, \citet{matt:2005} showed
the possibility that the stellar wind along the open magnetic field
originated from the star can cause significant spin-down torque on
the star, provided that mass-loss rate is high enough.

The main aim of this paper is to find a simple kinematic model which
can reproduce the wide variety of the observed profiles, and to perform
the empirical studies of the line formation. More specifically, we
present disc--wind--magnetosphere hybrid models for CTTS, and perform
parametric studies of the H$\alpha$ formation. This study should
provide preliminary physical conditions which lead to the wide variety
of emission line profiles seen in the observation, and will help
to construct more comprehensive circumstellar models (e.g. via MHD
simulations) of T~Tauri stars. Using the model results, we examine
the H$\alpha$ spectroscopic classification used by \citet{reipurth:1996},
and compare models with observations. We will also discuss whether
our model is consistent with some predictions made by the recent MHD
studies i.e.~ $\mu=\dot{M}_{\mathrm{wind}}/\dot{M}_{\mathrm{acc}}\approx0.1$
(e.g.~\citealt{koenigl:2000}).

In section~\ref{sec:model-configuration}, the model assumptions,
and the basic model configurations are presented. We discuss the radiative
transfer model used to compute the profiles in section~\ref{sec:radiative-transfer-model},
and the results of model calculations are given in section~\ref{sec:Results}.
The closer examination and discussion of the results are presented
in section~\ref{sec:Discussion}. Finally, the summary of this work
and the conclusion are in section~\ref{sec:Conclusions}.

@@

To understand how the different part of the CTTS circumstellar environment
contributes to the formation of H$\alpha$, the model space is divided
into four different regions: (1)~a central continuum source, (2)~the magnetospherical
accretion flow, (3)~the wind outflow, and (4)~the accretion
disc. Fig.~\ref{fig:config_dipolar} 
depicts the relative location of the regions in the model space. In
all regions, the density is symmetric around the z-axis. The innermost radius of
the magnetosphere at the equatorial plane coincides with the the inner
radius of the accretion disc. From the innermost part of the accretion
disc, the gas falls freely, moving along the magnetic field onto the
surface of the star. For simplicity,
the out-flowing wind component is defined only in the outside of the
largest radius of the magnetosphere (to avoid overlapping of the wind
with the magnetosphere). In the following subsections, the details
of model components will be described.
\begin{figure*}

@@

\caption{The basic model configuration. The system consists of four
components: (1)~the continuum source located at the origin $\left( o
\right)$ of the cartesian coordinates $\left(x,y,z\right)$ -- the
y-axis are into the paper, (2)~the magnetospherical accretion flow,
(3)~the bipolar wind outflow, and (4)~the accretion disc. The density
distribution is symmetric around the z-axis. The innermost radius of
the magnetosphere (at the equatorial plane) coincides with the the
inner radius of the accretion disc. From the innermost part of the
accretion disc, the gas falls freely, moving along the magnetic field
onto the surface of the star. The bipolar wind starts from just
outside of the largest radius 
($R_{\mathrm{mo}}$) of the magnetosphere (dotted line). }

@@

\subsection{Continuum Source}

@@

Unless specified otherwise, we adopt stellar parameters of a typical
classical T~Tauri star for the central continuum source, i.e. radius
($R_{*}$), mass ($M_{*}$), and effective temperature the photosphere
($T_{\mathrm{ph}}$) are $2\, R_{\sun}$, $0.5\, M_{\sun}$, and $4000\,\mathrm{K}$
respectively. The model atmosphere of \citet{kurucz:1993} with
$T_{\mathrm{ph}}=4000\,\mathrm{K}$ and $\log g_{*}=3.5$ (cgs) is used
to find the photospheric continuum at the H$\alpha$ wavelength. The
parameters are summarised in Table~\ref{tab:std_parameters}.

For a model which includes magnetospherical accretion, an additional
continuum source is considered. As the in-falling gas 
becomes close to the stellar surface, it is decelerates in a strong
shock, and is heated to $\sim 10^{6}\,\mathrm{K}$.
The X-ray radiation produced in the shock will be absorbed by the
gas locally, and re-emitted in optical and UV radiation (\citealt{koenigl:1991};
\citealt*{hartmann:1994}). This will create hot rings near the stellar
surface where the magnetic field intersects with the surface. For
simplicity, the free-falling kinetic energy is assumed to be thermalized
in the radiating layer, and re-emits as blackbody radiation (with
a single temperature). With the parameters of the magnetosphere and
the star given above (Table~\ref{tab:std_parameters}), about $8$~per~cent
of the surface is covered by the hot rings. If the mass-accretion rate is $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$,
the ratio of this accretion luminosity to the photospheric luminosity
is about $0.5$, and the corresponding temperature of the hot rings is
about $6400\,\mathrm{K}$.  The continuum emission from the hot rings is taken
into account when computing the line profiles.

@@

@@

We adopt the magnetospherical accretion flow model of \citet{hartmann:1994},
as done  by \citet{muzerolle:2001} and by \citet*{symington:2005},
in which the gas accretion on to the stellar surface from the innermost
part of the accretion disc occurs through a dipolar stellar magnetic
field. The magnetic field is assumed to be so strong that the gas
flow does not affect the underlying magnetic field itself. As shown
in Fig.~\ref{fig:config_dipolar}, the innermost radius ($R_{\mathrm{mi}}$)
of the magnetosphere at the equatorial plane ($z=0$) is assigned
to be same as the inner radius ($R_{\mathrm{di}}$) of the accretion
disc where the flow is truncated. In our models, $R_{\mathrm{mi}}$
and the outer radius ($R_{\mathrm{mo}}$) of the magnetosphere (at
the equatorial plane) are set to be $2.2\, R_{\sun}$ and $3.0\, R_{\sun}$
respectively. The former value corresponds to the co-rotation radius
of the accretion disc, and the geometry of the magnetic field/stream
lines is kept constant throughout this paper. The geometry of the
magnetosphere is identical to the {}``small/wide'' model of \citet{muzerolle:2001}.

The magnetic field and the gas stream lines are assumed to have the
following simple form: \begin{equation}
r=R_{\mathrm{m}}\sin^{2}\theta\label{eq:dipole}\end{equation}
 \citep*[see][]{ghosh:1977} where $r$, and $\theta$ are coordinates
of the field point ($p$) in Fig.~\ref{fig:config_dipolar} in spherical
coordinates, and $R_{\mathrm{m}}$ is the radial distance to the field
line at the equatorial plane ($\theta=\pi/2$). The range of
$R_{\mathrm{m}}$ is
restricted between $R_{\mathrm{mi}}$ and $R_{\mathrm{mo}}$. Using
the field geometry above and the conservation of energy, the velocity
and the density of the accreting gas along the steam line are found
as in \citet{hartmann:1994}.

The temperature structure of the magnetospheric used by \citet{hartmann:1994}
is simply adopted here. They computed the temperature assuming a volumetric
heating rate which is proportional to $r^{-3}$, and using the energy
balance of the radiative cooling rate of \citet{hartmann:1982} and
the heating rate \citep{hartmann:1994}. Although the temperature
structure of the accretion steam could significantly affect the line
source function, for the purpose of the exploring the general characteristics
of the H$\alpha$ formation, this simple form is a reasonable assumption.
\citet{martin:1996} presented the self-consistent determination of
the thermal structure of the in-flowing gas along the dipole magnetic
field (equation~\ref{eq:dipole}) by solving the heat equation couple
to rate equations for hydrogen. He found that main heat source is
adiabatic compression due to the converging nature of the flow, and
the major contributors to the cooling process are bremsstrahlung radiation
and line emission from Ca~II and Mg~II ions. The results of \citet{martin:1996}
qualitatively agree with that of \citet{hartmann:1994}.

\subsection{Outflow geometries}
Two different types of the wind configurations are considered: (1)~the
bipolar wind model which resembles the density structure of the MHD
simulation by \citet{krasnopolsky:2003} in \em large scale \em ($\sim
100$~au), and (2)~the disc-wind model which resembles the density of
the same MHD simulation in \em small scale \em ($\sim 10$~au).  While
the wind density is more concentrated toward the polar directions in
(1), it is more concentrated toward the equatorial direction in
(2). The following describes each model in details.

\subsubsection{Bipolar Wind}

@@

Similar to the simple wind model of \citet{appenzeller:2005}, the
following parametrisation of collimated bipolar wind (region 3 in
Fig.~\ref{fig:config_dipolar}) is adopted. The wind velocity field,
$\boldsymbol{v}_{\mathrm{wind}}$, consists of
radial and azimuthal components which depend on the spherical coordinates
$r$ and $\theta$. The radial component $v_{r}\left(r\right)$ is
assumed to be in the classical beta-velocity law \citep[c.f.][]{castor:1979},
and the azimuthal component $v_{\phi}$ is assumed to be a constant
fraction ($\gamma$) of the Keplerian velocity for a given distance
($w=\sqrt{x^{2}+y^{2}}$) from the symmetry axis ($z$-axis) , i.e.:

\begin{eqnarray}
\boldsymbol{v}_{\mathrm{wind}} & = & v_{r}\hat{\boldsymbol{r}}+v_{\phi}\hat{\boldsymbol{\bphi}}\label{eq:wind_velocity_def}\end{eqnarray}
where \begin{equation}
v_{r}\left(r\right)=v_{r0}+\left(v_{\infty}-v_{r0}\right)\left(1-\frac{R_{\mathrm{mo}}}{r}\right)^{\beta}\,,\label{eq:wind_velocity_radial_part}\end{equation}
\begin{equation}
v_{\phi}\left(w\right)=\gamma\,\left(\frac{GM_{*}}{w}\right)^{1/2}\,,\label{eq:wind_velocity_azimuth_part}\end{equation}
Note that the base of the wind starts at $r=R_{\mathrm{mo}}$. This
is chosen so mainly for computational simplicity i.e. to avoid the
overlap between the wind and the magnetosphere; however, see \citet{matt:2005}
for the possibility that wind originates from the stellar surface
(restricted near the polar caps). The range of the polar angle for
the wind is restricted to $\theta>\left|\theta_{\mathrm{disc}}\right|$
where $\theta_{\mathrm{disc}}$ is the opening angle of the accretion
disc, to avoid the overlap. $v_{r0}$ is the small radial velocity
at the base of the wind ($r=R_{\mathrm{mo}}$). Normally, $v_{r0}=10\,\mathrm{km\, s^{-1}}$
which is approximately equal to the thermal velocity of hydrogen with
$T=7500\,\mathrm{K}$. Following \citet{appenzeller:2005}, $\gamma=0.05$
is adopted. The dependency of $v_{r}$ in polar direction ($\theta=0$)
on the values of wind acceleration parameter $\beta$ is shown in
Figure~\ref{fig:wind_vr_rho}. All other parameters describing the
wind are fixed as the standard values given in Table~\ref{tab:std_parameters}.

Further, the density of the wind is assumed to be axi-symmetric and
separable in $r$ and $\theta$,

>>>>

We present  radiative transfer models of the circumstellar environment of 
classical T~Tauri stars, concentrating on the formation of the H$\alpha$ 
emission. The wide variety of line profiles seen in observations are
indicative of both inflow and outflow, and we therefore employ
a circumstellar structure that includes both magnetospheric accretion
and a wind. We adopt two kinematic models for the outflow, a generic
bipolar wind and a disc wind. We perform systematic investigations of
the model parameters for the wind and the magnetosphere to search
for possible geometrical and physical conditions which lead to the
types of profiles seen in observations. We find that both wind models
can reproduce the wide range profile types seen is observations, 
and that the most common profile types observed occupy a large
volume of parameter space. Conversely, the most infrequently observed
profile morphologies require a very specific set of models parameters.
Although a particular outflow model (bipolar or disc-wind) is not
well-constrained by a single line profile observations, we find that
the inclination dependence of the line equivalent width
predicted by the bipolar wind model agree with trends seen in the
observation, but the disc-wind model does not. Using the model results,
we examine the H$\alpha$ spectroscopic classification used by Reipurth
et.~al, and discuss the basic physical conditions that are required to reproduce
the profiles in each classified type.

@@

Many observational studies (e.g., \citealt{herbig:1962}; \citealt{edwards:1994};
\citealt{kenyon:1994}; \citealt*{reipurth:1996}; \citealt{alencar:2000})
of CTTS line profiles have revealed evidence for both outward wind flows and
inward accretion flows, characterized by the blue-shifted absorption features
in H$\alpha$ profiles and the redshifted inverse P~Cygni (IPC) profiles
respectively. Typical mass-loss rates of CTTS are about $10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$
to $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$ (e.g., \citealt{kuhi:1964};
\citealt{edwards:1987}; \citealt*{hartigan:1995}), and the mass-accretion
rates are also about $10^{-9}\,\mathrm{M_{\sun}\, yr^{-1}}$ to $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$
(e.g., \citealt{kenyon:1987}; \citealt*{bertout:1988}; \citealt{gullbring:1998}).
Recent H$\alpha$ spectro-astrometric observations by \citet*{takami:2003}
provide indirect evidence for the presence of bipolar and monopolar
outflows down to $\sim 1\,\mathrm{au}$ scale (e.g.\,CS~Cha and RU~Lup).
Similarly, ESO VLT observations using high-resolution ($R=50\,000$)
two-dimensional spectra of edge-on CTTS (HH30$^{*}$, HK~Tau~B,
and HV~Tau~C) by \citet{appenzeller:2005} show extended H$\alpha$
emission in the direction perpendicular to the obscuring circumstellar
disc,  suggesting the presence
of the bipolar outflows. On an even  larger scale, \emph{HST} observations
of HH30 \citep{burrows:1996} trace the jet to within $\lesssim30$~au of  the star. The jet has a cone shape with an opening angle of $3^{\circ}$
between 70 and 700 au \citep{koenigl:2000}. \citet{alencar:2000}
found about 80 per cent of their sample (30 CTTS) show blue-shifted
absorption components in at least one of the Balmer lines and Ca~K.

In the currently favoured model of accretion in CTTS, the
accretion disc is disrupted by the magnetosphere, which
channels the gas from the disc onto the stellar surface (e.g.,
\citealt{uchida:1985}; \citealt{koenigl:1991}; \citealt{cameron:1993};
\citealt{shu:1994}). This picture is supported
by recent measurements of strong ($\sim 10^{3}\mathrm{G}$)
magnetic fields in CTTS  (e.g., \citealt{johns-krull:1999};
\citealt{guenther:1996}xxxx;  \citealt{symington:2005b})
and by radiative transfer models which reproduce the gross characteristics of observed
profiles for some TTS (\citealt*{muzerolle:2001}). In particular the magnetospheric accretion (MA) model
explains blue-ward asymmetric emission line profiles as resulting from the partial occultation of
the flow by the  stellar photosphere, while inverse P~Cygni profiles in the MA model result from
inflowing material  at near free-fall velocities seen projected against hotspots on the stellar surface.

Despite these successes,  the overwhelming observational evidence for outflow
in the CTTS  suggests that the MA  model is only one component of a complex
circumstellar environment. Clearly, one must include the contribution of any
wind/jet flow if one wishes to both accurately predict the mass-accretion
rate and also determined the mass-loss rate of CTTS via emission profile
modelling. The first attempt in this direction was made by \citet{alencar:2005} who demonstrated that
 the observed  H$\alpha$, H$\beta$ and Na~D lines of RW Aur are better reproduced
by the radiative transfer model which included a collimated disc-wind
arising from near the inner edge of the accretion disc.

%Prior to the MA paradigm, many alternative models had
%been considered to explain the observed spectroscopic features mentioned
%above. For example, (1)~an Alfv\'{e}n wave-driven wind model
%(e.g.~\citealt{decampli:1981}; \citealt*{hartmann:1982}), (2)~a turbulent
%boundary layer (between the accretion disc and stellar surface) model
%(e.g.~\citealt{bertout:1988}; \citealt{basri:1989}), (3)~a chromospheric
%model (e.g.~\citealt*{calvet:1984}), and (4)~a disc wind model (e.g.
%\citealt*{calvet:1992b}; \citealt{kwan:1995}).

The main aim of this paper is to find a simple kinematic model which
can reproduce the wide variety of the observed profiles, and to perform
empirical studies of line formation in an attempt to place morphological
classification schemes on a firmer physical footing. We will also discuss whether
our model is consistent with some predictions made by recent MHD
studies i.e.~ $\mu=\dot{M}_{\mathrm{wind}}/\dot{M}_{\mathrm{acc}}\approx0.1$
(e.g.~\citealt{koenigl:2000}).

In section~\ref{sec:model-configuration}, the model assumptions,
and the basic model configurations are presented. We discuss the radiative
transfer model used to compute the profiles in section~\ref{sec:radiative-transfer-model},
and the results of model calculations are given in section~\ref{sec:Results}.
We discuss our results in the context of Reipurth's classification scheme in section~\ref{sec:Discussion}, and our summary and conclusions are presented in section~\ref{sec:Conclusions}.

@@

In order to understand how the different parts of the CTTS circumstellar environment
contribute to the formation of H$\alpha$, the model space is divided
into four different regions: (1)~a central continuum source, (2)~the magnetospheric
accretion flow, (3)~the wind outflow, and (4)~the accretion
disc. Fig.~\ref{fig:config_dipolar} 
depicts the relative location of the regions in the model space.
The density is assumed to be rotationally symmetric around the $z$-axis. The innermost radius of
the magnetosphere at the equatorial plane coincides with the the inner
radius of the accretion disc. From the innermost part of the accretion
disc, the gas falls freely, moving along the magnetic field onto the
surface of the star.  In the following subsections we describe the details
of model components.
\begin{figure*}

@@

\caption{The basic model configuration. The system consists of four
components: (1)~the continuum source located at the origin $\left( o
\right)$ of the cartesian coordinates $\left(x,y,z\right)$ -- the
$y$-axis is into the paper, (2)~the magnetospherical accretion flow,
(3)~the bipolar wind outflow, and (4)~the accretion disc. The density
distribution is symmetric around the $z$-axis. The innermost radius of
the magnetosphere (at the equatorial plane) coincides with the the
inner radius of the accretion disc. From the innermost part of the
accretion disc, the gas falls freely, moving along the magnetic field
onto the surface of the star. The bipolar wind starts from just
outside the largest radius 
($R_{\mathrm{mo}}$) of the magnetosphere (dotted line). }

@@

\subsection{The continuum source}

@@

We adopt stellar parameters of a typical
classical T~Tauri star for the central continuum source, i.e. radius
($R_{*}$), mass ($M_{*}$), and effective temperature the photosphere
($T_{\mathrm{ph}}$) are $2\, R_{\sun}$, $0.5\, M_{\sun}$, and $4000\,\mathrm{K}$
respectively. The model atmosphere of \citet{kurucz:1993} with
$T_{\mathrm{ph}}=4000\,\mathrm{K}$ and $\log g_{*}=3.5$ (cgs) defines
the photospheric contribution to the continuum flux. The
parameters are summarised in Table~\ref{tab:std_parameters}.

An additional continuum source is considered for models which includes 
magnetospheric accretion: as the infalling gas 
approaches the stellar surface, it decelerates in a strong
shock, and is heated to $\sim 10^{6}\,\mathrm{K}$.
The X-ray radiation produced in the shock will be absorbed by the
gas locally, and re-emitted in optical and UV radiation (\citealt{koenigl:1991};
\citealt*{hartmann:1994}). This will create hot rings  where the magnetic field
intersects with the surface. We assume that the free-falling kinetic energy is thermalized
in the radiating layer, and is re-emitted  as blackbody radiation with
a single temperature. With the parameters of the magnetosphere and
the star given above (Table~\ref{tab:std_parameters}), about $8$~per~cent
of the surface is covered by the hot rings. If the mass-accretion rate is $10^{-7}\,\mathrm{M_{\sun}\, yr^{-1}}$,
the ratio of this accretion luminosity to the photospheric luminosity
is about $0.5$, and the corresponding temperature of the hot rings is
about $6400\,\mathrm{K}$.  The continuum emission from the hot rings is taken
into account when computing the line profiles.

@@

\subsection{The magnetosphere}

@@

We adopt the MA flow model of \citet{hartmann:1994},
as done  by \citet{muzerolle:2001} and by \citet*{symington:2005},
in which the gas accretion on to the stellar surface from the innermost
part of the accretion disc occurs through a dipolar stellar magnetic
field. The magnetic field is assumed to be so strong that the gas
flow does not affect the underlying magnetic field itself. As shown
in Fig.~\ref{fig:config_dipolar}, the innermost radius ($R_{\mathrm{mi}}$)
of the magnetosphere at the equatorial plane ($z=0$) is assigned
to be same as the inner radius ($R_{\mathrm{di}}$) of the accretion
disc where the flow is truncated. In our models, $R_{\mathrm{mi}}$
and the outer radius ($R_{\mathrm{mo}}$) of the magnetosphere (at
the equatorial plane) are set to be $2.2\, R_{\sun}$ and $3.0\, R_{\sun}$
respectively. The former value is  the co-rotation radius
of the accretion disc, and the geometry of the magnetic field/stream
lines is fixed for all calculations. We note that this magnetospheric geometry
is identical to the {}``small/wide'' model of \citet{muzerolle:2001}.

The magnetic field and the gas stream lines are assumed to have the
following simple form: \begin{equation}
r=R_{\mathrm{m}}\sin^{2}\theta\label{eq:dipole}\end{equation}
 \citep*[see][]{ghosh:1977} where $r$, and $\theta$ are coordinates
of the field point ($p$) in Fig.~\ref{fig:config_dipolar} in spherical
coordinates, and $R_{\mathrm{m}}$ is the radial distance to the field
line at the equatorial plane ($\theta=\pi/2$). The range of
$R_{\mathrm{m}}$ is
restricted between $R_{\mathrm{mi}}$ and $R_{\mathrm{mo}}$. Using
the field geometry above and conservation of energy, the velocity
and the density of the accreting gas along the steam line are found
as in \citet{hartmann:1994}.

The temperature structure of the magnetospheric used by \citet{hartmann:1994}
is adopted here. They computed the temperature, assuming a volumetric
heating rate which is proportional to $r^{-3}$, by solving the energy
balance of the radiative cooling rate of \citet{hartmann:1982} and
the heating rate \citep{hartmann:1994}. 
\citet{martin:1996} presented a self-consistent determination of
the thermal structure of the inflowing gas along the dipole magnetic
field (equation~\ref{eq:dipole}) by solving the heat equation coupled
to the rate equations for hydrogen. He found that main heat source is
adiabatic compression due to the converging nature of the flow, and
the major contributors to the cooling process are bremsstrahlung radiation
and line emission from Ca~II and Mg~II ions. The results of \citet{martin:1996}
qualitatively agree with that of \citet{hartmann:1994}.

\subsection{Outflow geometries}

The magneto-centrifugal wind paradigm, first proposed by \citet{blandford:1982},
has been often used to model the large-scale wind structure of
T Tauri stars, or the observed optical jets (e.g.~HH~30 jet
by \citealt{burrows:1996}; \citealt{ray:1996}). The launching of
the wind from a Keplerian disc is typically done by treating the equatorial
plane of the disc as a mass-injecting boundary condition (e.g., \citealt{shu:1994};
\citealt{ustyugova:1995}; \citealt{ouyed:1997}; \citealt{krasnopolsky:2003}).
Depending on the location of the open magnetic fields anchored to
the disc, two different types of winds are produced. If the field
is constrained to be near the co-rotation radius of stellar magnetosphere,
an {}``X-wind'' \citep{shu:1994} is produces. If the open field
lines are located in a wider area of the disc, a {}``disc-wind''
similar to that of \citet{koenigl:2000} is produced \citep{krasnopolsky:2003}.
Recent reviews on the jet/wind-disc connection can be found in \citet{koenigl:2000}
and \citet{pudritz:2005}.

Clearly there are several alternative outflow scenarios. Here we adopt
two simple kinematical models which broadly represent the results of
MHD simulations. The first is a bipolar wind model, in which the outflow is collimated
along the rotational axes (a model similar to that of \citet{krasnopolsky:2003}). The
second model is based on the disc-wind paradigm, in which the density of the outflow
increases towards the equatorial plane. The  difference in latitudinal density
structure between the two outflow models should provide an indication as to whether
line profiles alone can be used to constrain the outflow geometry. In the following
sections we present details of the two outflow structures.

\subsubsection{The bipolar wind}

@@

In a similar fashion to the simple model of  \citet{appenzeller:2005}, the
following parametrisation of collimated bipolar wind (region 3 in
Fig.~\ref{fig:config_dipolar}) is adopted. The wind velocity field,
$\boldsymbol{v}_{\mathrm{wind}}$, consists of
radial and azimuthal components which depend on the spherical coordinates
$r$ and $\theta$. The radial component $v_{r}\left(r\right)$ is
assumed to follow the classical beta-velocity law \citep[c.f.][]{castor:1979},
and the azimuthal component $v_{\phi}$ is assumed to be a constant
fraction ($\gamma$) of the Keplerian velocity for a given distance
($w=\sqrt{x^{2}+y^{2}}$) from the symmetry axis ($z$-axis) , i.e.:

\begin{eqnarray}
\boldsymbol{v}_{\mathrm{wind}} & = & v_{r}\hat{\boldsymbol{r}}+v_{\phi}\hat{\boldsymbol{\bphi}}\label{eq:wind_velocity_def}\end{eqnarray}
where \begin{equation}
v_{r}\left(r\right)=v_{r0}+\left(v_{\infty}-v_{r0}\right)\left(1-\frac{R_{\mathrm{mo}}}{r}\right)^{\beta}\,,\label{eq:wind_velocity_radial_part}\end{equation}
\begin{equation}
v_{\phi}\left(w\right)=\gamma\,\left(\frac{GM_{*}}{w}\right)^{1/2}\,,\label{eq:wind_velocity_azimuth_part}\end{equation}
Note that the base of the wind starts at $r=R_{\mathrm{mo}}$. This naturally leads
to a density discontinuity, but allows a simple parameterization of the wind whereas
a self-consistent interface between the magnetosphere and wind would require
a full MHD calculation that is beyond the scope of this paper. The range of polar angle for
the wind is restricted to $\theta>\left|\theta_{\mathrm{disc}}\right|$
where $\theta_{\mathrm{disc}}$ is the opening angle of the accretion
disc. $v_{r0}$ is the small radial velocity
at the base of the wind ($r=R_{\mathrm{mo}}$). Normally, $v_{r0}=10\,\mathrm{km\, s^{-1}}$
which is approximately equal to the thermal velocity of hydrogen with
$T=7500\,\mathrm{K}$. Following \citet{appenzeller:2005}, $\gamma=0.05$
is adopted. The dependency of $v_{r}$ in polar direction ($\theta=0$)
on the values of wind acceleration parameter $\beta$ is shown in
Figure~\ref{fig:wind_vr_rho}. All other parameters describing the
wind are fixed as the standard values given in Table~\ref{tab:std_parameters}.

The density of the wind is assumed to be axi-symmetric and
separable in $r$ and $\theta$,

\begin{equation}
\rho\left(r,\theta\right)=P\left(r\right)\, F\left(\theta\right)\label{eq:wind_density_def}\end{equation}
 with \begin{equation}
F\left(\theta\right)=n\,\cos^{b}\theta\label{eq:wind_density_theta}\end{equation}
 where $b$ is normally a positive even number (for density symmetric
about the equatorial plane), and $n$ is the angular normalisation
constant. For $b=0$, the wind is spherically symmetric except for
the parts disrupted by the accretion disc. The larger the value of
$b$, the higher the degree of the collimation. By integrating equation~\ref{eq:wind_density_theta}
over angles and normalising the integral to $4\pi$, one finds \begin{equation}
n=\frac{1+b}{1-\cos^{1+b}\theta_{\mathrm{wind}}}\,.\label{eq:wind_density_theta_norm}\end{equation}
 Assuming the total mass-loss rate via the wind is $\dot{M}_{\mathrm{wind}}$
and the mass-flux is constant, the radial part of the density
function is reduced to $P\left(r\right)=\dot{M}_{\mathrm{wind}}\,\left[4\pi r^{2}v_{r}\left(r\right)\right]^{-1}$;
hence, equation~\ref{eq:wind_density_def} becomes \begin{equation}
\rho\left(r,\theta\right)=\frac{n\,\dot{M}_{\mathrm{wind}}\,\cos^{b}\theta}{4\pi r^{2}\, v_{r}\left(r\right)}\,.\label{eq:wind_density_final}\end{equation}
 For a given mass-accretion rate, the wind mass-loss rate in our typical
model is assigned from the ratio of mass-loss to mass-accretion rate
$\dot{M}_{\mathrm{wind}}/\dot{M}_{\mathrm{acc}}\approx0.1$, a value indicated
by both observations and MHD calculations (e.g. \citealt{koenigl:2000}).
Figure~\ref{fig:wind_vr_rho} shows the density along a streamline
in the polar direction ($\theta=0$), for different values of $\beta$
with all other parameters fixed at the standard values (Table~\ref{tab:std_parameters}).
The density is relatively sensitive to the value of $\beta$ for $r<10\, R_{\mathrm{mo}}$.
